Kindly Bent to Free Us

### 2019-06-03 kindly/motivation.tex
First shot at explaining imperative prog.

In addition, this function can be partially
applied to a channel. Since the closure returned by such a partial
application captures the channel, it can only be used once.
This is reflected by the arrow, \lstinline/-{lin}>/, which is the
type of functions that can only be used once.
The general form of arrows types in \lang is \lstinline/-{k}>/, where
\lstinline/kk/ is an arbitrary kind that describes the behavior of
underlying closure.
For convenience, we shorten \lstinline/-{un}>/ as \lstinline/->/.
Note again that the linearity on the arrows was inferred without
any annotation from the user. In fact, the only difference between the code
presented so far and the example by
\citet{DBLP:journals/jfp/Padovani17}
is the kind annotation on the type definition of \lstinline/st/.

@@

The programming style used for channels so far, which are passed back
and forth between functions, is appropriate for purely functional code.
This style is however fairly bothersome in imperative contexts.

ML languages are often characterised by their hybrid programming style
that mixes functional styles with imperative side effects. In
this context, the programming style used for channels is fairly verbose.
Similarly to Rust, we use {\em borrows} to allow easy imperative
programming with linear and affine resources.

>>>>

In addition, this function can be partially
applied to a channel. Since the closure returned by such a partial
application captures the channel, it can only be used once.
This is reflected by the arrow, \lstinline/-{lin}>/, which is the
type of functions that can only be used once.
The general form of arrows types in \lang is \lstinline/-{k}>/, where
\lstinline/kk/ is an arbitrary kind that describes the behavior of
underlying closure.
For convenience, we shorten \lstinline/-{un}>/ as \lstinline/->/.
Note again that the linearity on the arrows was inferred without
any annotation from the user. In fact, the only difference between the code
presented so far and the examples by
\citet{DBLP:journals/jfp/Padovani17}
is the kind annotation on the type definition of \lstinline/st/.

@@

The programming style used for channels so far is
appropriate for purely functional code.
This style is however fairly bothersome in mutable contexts.
%
To make linearity easier to use for imperative programming, we rely on
{\em borrows}. We present a typical API for mutable affine arrays
in \cref{sig:array} and an example of use in \cref{ex:array}.
We can distinguish three type of operations. Firstly,
function that manipulate the whole
array such as \lstinline/create/ and \lstinline/free/, and behaves exactly
like the functions on channel presented in the previous section.
Second, read-only functions such as \lstinline/get/
manipulates \emph{immutable} borrows. An immutable borrow over array \lstinline/a/
is introduced with \lstinline/&a/ and represent a read-only view of the array.
Finally, mutable functions, such as \lstinline/set/ manipulate
a \emph{mutable borrow}. Mutable borrows, introduced with \lstinline/&!a/
represent a read-write view of the array.

Multiple immutable borrows can be used concurrently and are considered
unrestricted (\lstinline/un/).
A mutable borrow is exclusive, and considered affine (\lstinline/aff/).
Furthermore, any borrow prevents access to the ``raw'' access
as long as they are in use. In \cref{ex:array}, on \cref{ex:array:get},
two immutable borrows are used in the same expression. Once this operation
is done, the borrows are not used anymore, which allows us to take
an exclusive immutable borrow to mutate the array on \cref{ex:array:set}.
Finally we can return the whole array on \cref{ex:array:return} since
none of the borrows are used anymore.

\begin{figure}
  \centering
  \begin{subfigure}{0.48\linewidth}
\begin{lstlisting}
module Array : sig
  type ('a : un) t : aff

  val create : ('a : un) => int -> 'a -> 'a t
  val free : 'a t -> unit

  val length : &('a t) -> int

  val get : &('a t) * int -> 'a
  val set : &!('a t) * int * 'a -> unit
end
\end{lstlisting}
    \caption{Affine primitives}
    \label{sig:array}
  \end{subfigure}\hfill
  \begin{subfigure}{0.5\linewidth}
\begin{lstlisting}
let mk_fib_array n =
  let a = create n 1 in
  for i = 2 to n - 1 do
    let x1 = get &a (i-1) + get &a (i-2) in(*@\label{ex:array:get}*)
    set &!a i(*@\label{ex:array:set}*)
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}
    \caption{Example of use of affine arrays}
    \label{ex:array}
  \end{subfigure}
\end{figure}

\begin{figure}
  \centering
  \begin{subfigure}{0.5\linewidth}
\begin{lstlisting}
let for f a  = 
  let n = Array.length &a in
  let rec loop f a i = 
    if i < n then (
      f &!a i;
      loop f &!a (i+1)
    )
  in
  loop f &!a 0
# Borrowed m ('a Array.t) tt =>
  (tt -> int -> unit) -> tt -> unit
\end{lstlisting}
    \caption{For loop function for affine arrays}
    \label{ex:array}
  \end{subfigure}
\end{figure}
